Translation of B Implementations to the LLVM: Control and Modularity

### 2013-10-22 translation.tex
Compilation of translation.tex
* Put the required B-project.pdf (used convert B-project.tiff B-project.pdf)
* Fixed a very minor bug (description -> enumerate) in the tex file
Several bug fixes, including in the specification of the translation for operation calls. The translator now produces LLVM code for counter and wd, the latter importing the former. The produced LLVM code compiles without error with "llvm-as-mp-3.1".

Our approach consists in dividing the translation into three sections, and to
combine these sections differently in \emph{PROJ\/} and \emph{COMP\/} mode. For
a given module \B{M}, the sections are the following:
\begin{enumerate}
\item typedef: Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
  the data space of the module:
  \begin{tabbing}
    \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}}
  \end{tabbing}
  In case the data space of a module is empty, this section is empty and the
  type is not defined.
\item interface: Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data} (if
  defined):
  \begin{tabbing}
    \llvm{\%M\$ref\$ = type \%M\$state\$*}
  \end{tabbing}
  The interface also contains declarations of one LLVM functions for each
  operation of the module. 
  \begin{tabbing}
    \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})}
  \end{tabbing}
  If the data space of the module is not empty, a function definition
  corresponding to the initialisation is also emitted:
  \begin{tabbing}
    \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})}
  \end{tabbing}
\item implementation: Defines the functions declared in the interface:
  \begin{tabbing}
    foo \= \kill
    \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
    \> \llvm{\ListOf{\nt{block}}} \\
    \> \llvm{exit: ret void} \\
    \llvm{\}}
  \end{tabbing}
  This section is produced only for the developed modules.
\end{description}
Based on the definition of these sections, the \emph{COMP\/} translation
assembles them as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each imported module} \B{Q} \textit{generate} \\
    \> \llvm{\%Q\$state\$ = type opaque} \quad \textit{(declares type for} \B{Q} \textit{state space)} \\
    \> \textit{include the interface section of }\B{Q} \\
    \textit{include the typedef section of} \B{M} \\
    \textit{include the implementation section of} \B{M}
  \end{tabbing}
\end{center}
For a module \B{M}, the \emph{PROJ\/} translation is organized as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each transitively imported component} \B{Q} \\
    \> \textit{include the typedef section of \B{Q}} \\
    \textit{for each instance} \B{Q} \textit{imported transitively through} \B{path} \\
    \> \textit{declare a variable of type} \llvm{\%Q\$state\$}: \\
    \> \llvm{@Q$\lbrack$path$\rbrack$ = common global \%M\$state\$ zeroinitializer} \\

@@

\paragraph{Call up} instructions apply B operations with some values and obtain
results. The LLVM code generator produces instructions to evaluate the operation
arguments, the address of the variables receiving the results as well as an
instance of the LLVM \llvm{call} instruction. In addition, the proposed
translation includes as argument the address of a structure storing the state of
the component corresponding to this operation. Note that this component may be
either the implementation being translated (in the case of a so-called local
operation), or an instance of an imported machine.
\begin{align*}
  \trad{c}{\B{Call}} \isdef 
  & \LET p_i \sep il = \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol = 
  \trad{c.out}{\ListOf{Out}} \AND \\
  & \quad id = \Global(\B{c}) \AND ty = \llvm{"\PH{\state(\B{c.root})} *} \AND \\
  & \quad val \sep p = \IF \B{c.inst} = \text{None} \\
  & \quad \quad \THEN \llvm{"\%self\$"} \sep \llvm{""} \\
  & \quad \quad \ELSE \LET v = \variable \IN \\
  & \quad \quad \quad v \sep \\
  & \quad \quad \quad \llvm{"\PH{v} = getelementptr \PH{ \state({\B{c.inst.root}})} \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})} \nl"} \\
  & \IN \\
  & \quad \llvm{"\PH{p_i}} \\
  & \quad \llvm{\PH{p_o}} \\
  & \quad \llvm{\PH{p}} \\
  & \quad \llvm{call void \PH{id}(\PH{ty}* \PH{val}, \PH{il}, \PH{ol}) \nl"}
\end{align*}
The translation of input and output arguments produces two code snippets. The
first corresponds to the LLVM instruction sequence $p$ responsible for
evaluating the argument, and the second to the type $t$ and name $v$ of the
variable storing the resulting value:
\begin{align*}
  \trad{i}{Inp} \isdef 
  & \LET p \sep v \sep t = \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
  \trad{o}{Out} \isdef 
  & \LET p \sep v \sep t = \trad{o}{lv} \IN p \sep \llvm{"\PH{t} \PH{v}"}
\end{align*}

>>>>

Our approach consists in dividing the translation into three sections, and to
combine these sections differently in \emph{PROJ\/} and \emph{COMP\/} mode. For
a given module \B{M}, the sections are the following:
\begin{enumerate}
\item typedef: Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
  the data space of the module:
  \begin{tabbing}
    \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}}
  \end{tabbing}
  In case the data space of a module is empty, this section is empty and the
  type is not defined.
\item interface: Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data} (if
  defined):
  \begin{tabbing}
    \llvm{\%M\$ref\$ = type \%M\$state\$*}
  \end{tabbing}
  The interface also contains declarations of one LLVM functions for each
  operation of the module. 
  \begin{tabbing}
    \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})}
  \end{tabbing}
  If the data space of the module is not empty, a function definition
  corresponding to the initialisation is also emitted:
  \begin{tabbing}
    \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})}
  \end{tabbing}
\item implementation: Defines the functions declared in the interface:
  \begin{tabbing}
    foo \= \kill
    \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
    \> \llvm{\ListOf{\nt{block}}} \\
    \> \llvm{exit: ret void} \\
    \llvm{\}}
  \end{tabbing}
  This section is produced only for the developed modules.
\end{enumerate}
Based on the definition of these sections, the \emph{COMP\/} translation
assembles them as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each imported module} \B{Q} \textit{generate} \\
    \> \llvm{\%Q\$state\$ = type opaque} \quad \textit{(declares type for} \B{Q} \textit{state space)} \\
    \> \textit{include the interface section of }\B{Q} \\
    \textit{include the typedef section of} \B{M} \\
    \textit{include the implementation section of} \B{M}
  \end{tabbing}
\end{center}
For a module \B{M}, the \emph{PROJ\/} translation is organized as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each transitively imported component} \B{Q} \\
    \> \textit{include the typedef section of \B{Q}} \\
    \textit{for each instance} \B{Q} \textit{imported transitively through} \B{path} \\
    \> \textit{declare a variable of type} \llvm{\%Q\$state\$}: \\
    \> \llvm{@Q$\lbrack$path$\rbrack$ = common global \%M\$state\$ zeroinitializer} \\

@@

\paragraph{Call up} instructions apply B operations with some values and obtain
results. The LLVM code generator produces instructions to evaluate the operation
arguments, the address of the variables receiving the results as well as an
instance of the LLVM \llvm{call} instruction. In addition, the proposed
translation includes as argument the address of a structure storing the state of
the component corresponding to this operation. Note that this component may be
either the implementation being translated (in the case of a so-called local
operation), or an instance of an imported machine.
\begin{align*}
  \trad{c}{\B{Call}} \isdef 
  & \LET p_i \sep il = \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol = 
  \trad{c.out}{\ListOf{Out}} \AND \\
  & \quad id = {\Global(\B{c})} \AND ty = \llvm{"\PH{\stateref(\B{c.root})}} \AND \\
  & \quad val \sep p = \IF \B{c.inst} = \text{None} \\
  & \quad \quad \THEN \llvm{"\%self\$"} \sep \llvm{""} \\
  & \quad \quad \ELSE \LET v_1, v_2 = \variable, \variable \IN \\
  & \quad \quad \quad v_2 \sep \\
  & \quad \quad \quad \llvm{"\PH{v_1} = getelementptr \PH{ \state({\B{c.inst.root}})} \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})} \nl} \\
  & \quad \quad \quad \quad \llvm{\PH{v_2} = load \PH{ty}* \PH{v_1} \nl"} \\
  & \IN \\
  & \quad \llvm{"\PH{p_i}} \\
  & \quad \llvm{\PH{p_o}} \\
  & \quad \llvm{\PH{p}} \\
  & \quad \llvm{call void \PH{id}(\PH{ty} \PH{val}, \PH{il}, \PH{ol}) \nl"}
\end{align*}
The translation of input and output arguments produces two code snippets. The
first corresponds to the LLVM instruction sequence $p$ responsible for
evaluating the argument, and the second to the type $t$ and name $v$ of the
variable storing the resulting value:
\begin{align*}
  \trad{i}{Inp} \isdef 
  & \LET p \sep v \sep t = \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
  \trad{o}{Out} \isdef 
  & \LET p \sep v \sep t = \trad{o}{lv} \IN p \sep \llvm{"\PH{t} \PH{v}"}
\end{align*}
